---
title: "Creating, accessing, transforming, and manipulating dataframes"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    theme: cerulean

---

```{r include=FALSE}
# To see the output from this file - click on the Knit button 
```


```{r}

# Load knitr to allow us to print "pretty" tables
if(!require("knitr"))
   install.packages("knitr")
library(knitr)

```

# Creating a dataframe 

```{r}
animation_df <- data.frame(
  Movie = c("Shrek", "Up", "Jungle Book", "Aladdin"),
  Director = c("John Doe", "Jane Smith", "Alice Brown", "Bob White"),
  ReleaseYear = c(2020, 2021, 2019, 2022),
  BudgetMillions = c(150, 200, 180, 210),
  BoxOfficeMillions = c(500, 650, 550, NA)
)

print("Initial dataframe:")
print(animation_df)
```

# Accessing elements of a dataframe (and printing using Knitr::Kable)

```{r}

print("Accessing the 'Movie' column:")
kable(animation_df$Movie)

print("Accessing the first row:")
kable(animation_df[1,])
```

# Adding and removing columns 

```{r}

animation_df$DurationMinutes <- c(90, 100, 95, 105)
print("After adding 'DurationMinutes' column:")
kable(animation_df)

animation_df$DurationMinutes <- NULL
print("After removing 'DurationMinutes' column:")
kable(animation_df)

kable(animation_df, "html", caption = "The table with a column removed") 
```

# Adding and removing rows 

```{r}

new_movie <- data.frame(Movie="Virtual Voyage", Director="Gary Green", ReleaseYear=2023, BudgetMillions=220, BoxOfficeMillions=NA)
animation_df <- rbind(animation_df, new_movie)
print("After adding a new movie:")
kable(animation_df)

animation_df <- animation_df[-5,]
print("After removing the last movie:")
kable(animation_df)
```

# Basic operations on dataframe columns

```{r}

animation_df$ProfitMillions <- animation_df$BoxOfficeMillions - animation_df$BudgetMillions
print("After calculating profit for each movie:")
kable(animation_df)
```

# Filtering rows based on conditions 

```{r}

movies_after_2020 <- animation_df[animation_df$ReleaseYear > 2020,]
print("Movies released after 2020:")
kable(movies_after_2020)
```

# Sorting a dataframe 

```{r}

sorted_by_profit <- animation_df[order(animation_df$ProfitMillions, decreasing = TRUE), ]
print("Movies sorted by profit in descending order:")
kable(sorted_by_profit)
```

# Applying a function to dataframe columns 

```{r}

animation_df$ProfitabilityRatio <- sapply(animation_df$ProfitMillions, function(x) if(!is.na(x)) x/100 else NA)
print("After calculating profitability ratio (Profit/100):")
kable(animation_df)
```

# Merging dataframes 

```{r}

# Notice that the movies are OUT OF SEQUENCE - this is one of the clever aspects of the merge() function
ratings_df <- data.frame(
  Movie = c("Jungle Book", "Shrek",  "Aladdin", "Up"),
  IMDBRating = c(7.5, 8.3, 7.0, 8.0)
)

merged_df <- merge(animation_df, ratings_df, by="Movie")
print("After merging with ratings dataframe:")
kable(merged_df)
```

# Aggregating data using aggregate() 

```{r}

avg_budget_by_year <- aggregate(BudgetMillions ~ ReleaseYear, data=animation_df, FUN=mean)
print("Average budget by release year:")
kable(avg_budget_by_year)
```


# Using subset() for filtering 

```{r}

high_rated_movies <- subset(merged_df, IMDBRating >= 8.0)
print("Highly rated movies (IMDB Rating >= 8.0):")
kable(high_rated_movies)
```

# Using transform() for modifying columns 

```{r}

merged_df <- transform(merged_df, BudgetInBillions = BudgetMillions / 1000)
print("Budget in billions:")
kable(merged_df)
```

# Finding unique values with unique() 

```{r}

unique_directors <- unique(animation_df$Director)
print("Unique directors in the dataframe:")
kable(unique_directors)
```

# Getting summary statistics with summary() 

```{r}

print("Summary statistics:")
print(summary(animation_df))
```

# Handling missing values with na.omit() and is.na() 

```{r}

print("Movies with missing box office data:")
kable(animation_df[is.na(animation_df$BoxOfficeMillions),])

cleaned_df <- na.omit(animation_df)
print("Dataframe after omitting rows with missing values:")
kable(cleaned_df)
```

# Finding column-wise maximums and minimums 

```{r}


max_budget <- max(animation_df$BudgetMillions, na.rm = TRUE)
min_budget <- min(animation_df$BudgetMillions, na.rm = TRUE)
print(paste("Max Budget:", max_budget))
print(paste("Min Budget:", min_budget))
```

# Using duplicated() to find duplicate rows 

```{r}

animation_df$duplicated_row <- duplicated(animation_df)
print("Dataframe with duplicated row information:")
kable(animation_df)
```

# Using t() to transpose the dataframe 

```{r}

transposed_df <- as.data.frame(t(animation_df[1, -ncol(animation_df)])) # Transpose the first row (excluding the 'duplicated_row' column)
print("Transposed first row:")
kable(transposed_df)
```

# Group-wise operations using tapply() 

```{r}

avg_budget_by_year <- tapply(animation_df$BudgetMillions, animation_df$ReleaseYear, mean)
print("Average budget by release year using tapply():")
kable(avg_budget_by_year)
```

# Using split() to split data by factors 

```{r}

split_by_year <- split(animation_df, animation_df$ReleaseYear)
print("Dataframe split by release year:")
print(split_by_year)

```
