---
title: "Testing for Normality"
author: "Niall McGuinness"
date: "2023-11-17"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    theme: cerulean
---


```{r Load Packages, echo=FALSE}
# Load knitr to allow us to ask for user-input
if(!require("knitr"))
   install.packages("knitr")
library(knitr)
```


## Generate some random data with a Normal Distribtion

```{r}

# Set descriptive statistics for our vector of high scores
N <- 200
mn <- 210
std_dev <- 20

# Randomize the high scores using system time
set.seed(Sys.time())

# Generate a random Normal distribution of high scores with 99% of players in the range [150, 270] (i.e. 6 sigma or -3xSD on left and +3xSD on right of the mean)
high_scores <-  rnorm(N, mean = mn, sd = std_dev)

# Try using an Exponential distribution instead of a random normal distribution (see that Shapiro-Wilk no longer reports normality)
# high_scores <- rexp(N, rate = 0.1)

# Try using a Poisson distribution instead of a random normal distribution (see that Shapiro-Wilk no longer reports normality)
# high_scores <- rpois(N, lambda = 5) 

# Print scores
high_scores 

```

## Test for Normality


### Visual inspection - Histogram

```{r}
hist(high_scores)
# comment - modality, symmetry, skewness
```

### Visual inspection - QQ-Plot 

See https://www.statisticshowto.com/q-q-plots/

```{r}
qqnorm(high_scores)
qqline(high_scores)
```

### Descriptive Statistic - Shapiro-Wilk

The null hypothesis of the Shapiro-Wilk test is that the population is **normally** distributed. If the p-value associated with the test is below a chosen significance level (e.g., 0.05), then the null hypothesis is rejected, suggesting that the data does not follow a normal distribution.

```{r}
result_sw <- shapiro.test(high_scores)
result_sw
p_value <- result_sw$p.value
is_normal <- p_value > 0.05
```

A `r result_sw$method` was conducted on the data. From the output obtained we `r if(is_normal){"can"}else{"cannot"}` assume normality and **accept the null hypothesis** as the p-value (p=`r round(p_value, 3)`) is `r if(is_normal){"greater"}else{"less"}` than 0.05.


### Descriptive Statistic - Anderson-Darling


```{r}
if(!require("nortest"))
  install.packages("nortest")
library(nortest)

result_ad <- ad.test(high_scores)
result_ad
p_value <- result_ad$p.value
is_normal <- p_value > 0.05
```


An `r result_ad$method` was conducted on the data. From the output obtained we `r if(is_normal){"can"}else{"cannot"}` assume normality and **accept the null hypothesis** as the p-value (p=`r round(p_value, 3)`) is `r if(is_normal){"greater"}else{"less"}` than 0.05.
